Multitask matrix completion for learning protein interactions across diseases

### 2016-08-31 paper.tex
old bioinfo template

\documentclass[runningheads,a4paper]{llncs}
\usepackage[top=1.3in, bottom=1.3in, left=1in, right=1in]{geometry}
\usepackage{floatrow}
\usepackage{graphicx} % more modern
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{subfigure} 
\usepackage{natbib}
\usepackage[footnotesize]{caption}
\usepackage{multirow}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\newfloatcommand{capbtabbox}{table}[][11cm]

@@

\usepackage{url}
\urldef{\mailsa}\path|{mkshirsa,jgc,judithks,kmuruges}@cs.cmu.edu|
%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\argmin}{\operatornamewithlimits{argmin}}

\setcounter{tocdepth}{3}

\setcounter{secnumdepth}{0}

\begin{document}

\mainmatter

\title{Multitask matrix completion for learning protein interactions across diseases}

\titlerunning{Multitask matrix completion}

\author{Meghana Kshirsagar$^{1}$%
\and Jaime G. Carbonell$^{1}$
\and Judith Klein-Seetharaman$^{2}$
\and Keerthiram Murugesan$^{1}$}

\authorrunning{Multitask matrix completion}
% (feature abused for this document to repeat the title also on left hand pages)

\institute{$^{1}$Language Technologies Institute, Carnegie Mellon Univ., 5000 Forbes Ave., Pittsburgh PA 15213, USA\\
$^{2}$Metabolic \& Vascular Health, Warwick Medical School, Univ. of Warwick, Coventry, UK\\
\mailsa
}

\toctitle{Matrix completion for transfer learning in host-pathogen PPI networks}
\tocauthor{Meghana Kshirsagar, Jaime G. Carbonell, Judith Klein-Seetharaman and Keerthiram Murugesan}
\maketitle

\begin{abstract}

Disease causing pathogens such as viruses, introduce their proteins into the host cells 
where they interact with the host's proteins enabling the virus to replicate inside the host.
These interactions between pathogen and host proteins are key to understanding infectious diseases.
Often multiple diseases involve phylogenetically related or biologically similar pathogens. Here we present a multitask learning method to jointly model interactions between human proteins and three different, but related viruses: \textit{Hepatitis C}, \textit{Ebola} virus and \textit{Influenza A}. 
Our multitask matrix completion based model uses a shared low-rank structure in addition to a task-specific sparse structure to incorporate the various interactions.
We obtain between 7 to 39 \% points improvement in predictive performance over prior state-of-the-art models. 
We show how our model's parameters can be interpreted to reveal both general and specific
interaction-relevant characteristics of the viruses. Our code and data is available at: \url{http://www.cs.cmu.edu/~mkshirsa/bsl_mtl.tgz}
\keywords{host-pathogen protein interactions, multitask learning, matrix completion}
\end{abstract}

@@

This leads us to the following model that incorporates the above ideas. The interactions matrix $M_t$ of task $t$ can be written as: $M_t = \mu_t *\; \textrm{(shared component)} + (1-\mu_t) *\; \textrm{(specific component)} $, with hyperparameter $\mu_t$ allowing each task to customize its' amount of shared and specific components.

\begin{figure*}
\begin{floatrow}
\ffigbox[0.29\textwidth]{%
%trim option's parameter order: left bottom right top
\includegraphics[scale=0.33, trim = 0.2 9.2cm 4cm 0]{fig1.eps}
}
{
\caption{Multiple bipartite graphs: on the left are host proteins and on the right are proteins from different virus species. Edges represent protein interactions.}
\label{fig:hpppigraph}
}
\capbtabbox[0.69\textwidth]{%
\begin{footnotesize}
\begin{tabular}{c|r|r|r} \toprule
Task $\rightarrow$ &{\textit{Influenza A}} &\textit{Hepatitis C} & {\textit{Ebola}}\\
& (28 strains) & (5 strains,  & (2 strains) \\
& & many isolates) & \\ \midrule 
number of HP PPIs (positives) & 848& 981 & 90 \\\hline
number of non-homologous PPIs & 334 & 329 & 49 \\ \hline
\# of unique virus proteins in dataset & 54 & 151 & 2 \\\hline
\# of unique human proteins in dataset & 362 & 385 & 88 \\\hline
total \# of virus proteins in & 542 & 163 & 38 \\
UniprotKB across strains $^\dagger$ & & & \\ \hline
number of negatives & 84800 & 98100 & 9000 \\\hline
density of observed graph$^\ddagger$ (as \%) & .15 & .60 & .20 \\\hline
\multicolumn{4}{l}{\scriptsize{HP PPI: host-pathogen protein protein interactions}}\\
\multicolumn{4}{l}{\scriptsize{$\ddagger$ Density = (no. of positives + no. of negatives)/ total number of possible edges. The}} \\
\multicolumn{4}{l}{\scriptsize{\hspace{0.1cm} denominator was computed considering all proteins from all strains of the virus and}}\\
\multicolumn{4}{l}{\scriptsize{\hspace{0.1cm} $\approx$ 100,000 human proteins.}}\\
\multicolumn{4}{l}{\scriptsize{$\dagger$ this is the number of `reviewed' proteins on UniprotKB from all strains in our data}} \\
\bottomrule
\end{tabular}
\end{footnotesize}
}
{
\caption{Tasks and their sizes. Each column corresponds to one bipartite graph between human proteins and the pathogen indicated in the column header. All pathogens are single stranded RNA viruses. The last row shows that each of our graphs is extremely sparse.} 
\label{datasets}
}
\end{floatrow}
\end{figure*}

To incorporate the above ideas, we assume that the interactions matrix $M$ is generated from two components. 
The first component has low-rank latent factors over the human and virus proteins,
with these latent factors jointly learned over all tasks. The second component involves a task specific parameter, on which we
additionally impose a sparsity constraint as we do not want this parameter to overfit the data. Section \ref{ourmodel} discusses
our model in detail. We trade-off the relative importance of the two components using task-specific hyperparameters.
Our model can thus learn what is conserved and what is different across pathogens, rather than having to specify it manually.

@@

\noindent\textbf{Prediction of host-pathogen PPIs}: The most reliable experimental methods for studying protein-protein interactions (PPI) are often very time-consuming and expensive, making it hard to investigate the prohibitively large set of possible host-pathogen interactions -- for example, the bacterium \textit{Bacillus anthracis} which causes anthrax has about 2321 proteins which when coupled with the 100,000 or so human proteins gives $\approx$232 million protein pairs to test, experimentally. Computational techniques complement laboratory-based methods by predicting highly probable PPIs. These techniques use the known interactions data from previous experiments and predict the most plausible new interactions. In particular, supervised machine learning based methods use the few known interactions as training data and formulate the interaction prediction problem in a classification setting, with target classes: ``interacting'' or ``non-interacting''. Features are derived using various attributes of the two proteins such as: protein sequences from Uniprot \cite{uniprot}, protein structure from PDB, gene ontology from GO database \cite{GO}.

@@

@@

As before, the shared factors $U$ and $V$ are both $\mathbb{R}^{d_t \times k}$ (where the common dimensionality $d_t$ of the two node types is assumed for convenience). The matrix $S_t \in \mathbb{R}^{d_t \times d_t}$ is a sparse matrix. The objective function for the multitask model is given by:
\begin{equation}
\label{mtl_objective}
\begin{array}{ll}
\mathcal{L}(U,V,\{S_t\}) & = \displaystyle{ \frac{1}{N} \sum_{t=1}^T \sum_{(i,j) \in \Omega_t}} c^t_{ij} \; \ell \big( M_{t_{ij}}, \mathbf{x}_{ti}^\intercal H_t \mathbf{y}_{tj} \big)^2 + \lambda ( \|U\|^2_F + \|V\|^2_F ) + \displaystyle{\sum_{t=1}^T \sigma_t \|S_t\|_1} 
\end{array}
\end{equation}

@@

%\begin{algorithm}[h]   \caption{Alternating Least Squares}
%   \label{alg}
%   \begin{small}
%\begin{algorithmic}[1]
%\STATE \textbf{Input}: \\
%$k$ : number of latent factors\\
%$\Gamma$: pairs of entities for initialization\\
%For every task $t$,\\
%${\{\mathbf{x}_{ti}\}, \{\mathbf{y}_{tj}\}}$: feature vectors for pathogen and host proteins resp.\\
%$\Omega_t$: the observed entries of the matrix $M_t$\\
%\STATE \textbf{Initialization}: \\
%\STATE An iteration $r$, let $\mathbf{S}^r$ represent $\{S^r_t\}_{t=1}^T$
%\STATE $\mathbf{S}^0 \leftarrow 0$
%\STATE $U^0 \leftarrow$ top-$k$ left singular vectors and $V^0 \leftarrow$ top-$k$ right singular vectors from the SVD of $\displaystyle{\sum_{(i,j) \in \Gamma}} \mathbf{x}_i \mathbf{y}_j^\intercal$\\
%\STATE $\mathcal{L}^0$ : initial loss
%\REPEAT
%  \STATE $U^{r+1} \leftarrow \displaystyle{\argmin_{U}} \; \mathcal{L}(U, V^r, \mathbf{S}^r)$  \STATE $V^{r+1} \leftarrow \displaystyle{\argmin_{V}} \; \mathcal{L}(U^{r+1}, V, \mathbf{S}^r) $  \STATE For each task $t$\\
%  \hspace{0.5cm} $S^{r+1}_t \leftarrow \displaystyle{\argmin_{S_t}} \; \mathcal{L}(U^{r+1}, V^{r+1}, \mathbf{S}^r_{-t}) $
%  \STATE Compute $\mathcal{L}^{r+1}$ and let $\delta \leftarrow (\mathcal{L}^r - \mathcal{L}^{r+1})/\mathcal{L}^r$
%\UNTIL {$\delta < \tau$}%\While
%\end{algorithmic}\end{small}\end{algorithm}

\noindent \textbf{Optimization}:
The function $\mathcal{L}(U, V, \{S_t\})$ is non-convex. However, it is convex in every one of the parameters (i.e when the other parameters are fixed) 
and a block coordinate descent method called alternating least squares (ALS) is commonly used to optimize such functions. 
To speed up convergence we use an adaptive step size. %The detailed optimization procedure is shown in Algorithm \ref{alg}.

@@

\section{Experimental setup}
\subsection{Datasets and features}
\label{sec:datasets}
We use three human-virus PPI datasets from the PHISTO \citep{phisto} database (version from 2014), the characteristics of which are summarized in Table \ref{datasets}. 
The \textit{Influenza A} task includes various strains of flu: H1N1, H3N2, H5N1, H7N3. 
Similarly, the \textit{Hepatitis} task includes various strains of the virus. 
All three are single-strand RNA viruses, with \textit{Hepatitis} being a positive-strand ssRNA whereas \textit{Influenza} and \textit{Ebola} are negative-strand viruses. Note that we use UniprotKB identifiers to refer to
the proteins; hence if an identical protein appears in two strains it will be treated differently as it has a different
UniprotKB id. The density of the known interactions is quite small when considering the entire proteome (i.e all known proteins) of the host and pathogen species (row-4 in Table \ref{datasets}). \\
\noindent\textbf{Features}: 
Since the sequence of a protein determines its structure and consequently its function, it may be possible to predict PPIs using the amino acid sequence of a protein pair. \cite{shen:2007} introduced the ``conjoint triad model'' for predicting PPIs using only amino acid sequences. They partitioned the twenty amino acids into seven classes based on their electrostatic and water affinities.\footnote{For details of these classes, please refer to the supplementary or the original paper}
A protein's amino acid sequence is first transformed to a class-sequence (by replacing each amino acid by its class).
For $k$=3, they count the number of times each distinct tri-mer (set of three consecutive amino acids) occurred in the sequence.
Since there are 343 ($7^3$) possible tri-mers (with an alphabet of size 7), the feature vector containing the tri-mer frequency counts will have 343 elements.
To account for protein size, they normalized the counts by linearly transforming them to lie between 0 and 1.
Thus the value of each feature in the feature vector is the normalized count for each of the possible amino acid three-mers. We use di-, tri- and four-mers thus leading to a total of 2793 features ($7^2 + 7^3 + 7^4$).
Such features have been successfully applied in prior work \citep{dyer07,me_ismb_2013}.

@@

\subsection{Comparison with a homolog baseline}
In addition to the machine learning methods presented in Section \ref{mlmethods}, we also tried a simple homology based approach that is based on protein sequence identity. We use the entire human-virus interactome from PHISTO as input to this method. Given a human protein $h$, we rank all virus proteins $v$ based on the following simple scheme. let $V(h)$ be a set of all virus proteins interacting with the human protein $h$ and let $H(v)$ be all human proteins interacting with the virus protein $v$. We score each pair $(h, v)$ from the bipartite graph as follows: $\max \{ \max $ sequence identity between $h$ and all host proteins in $H(v)$; $\max $ sequence identity between $ v $ and all virus proteins in $V(h) \}$.

@@

\begin{table*}[t]\caption{Area Under the Precision-Recall curve for each task on the PHISTO dataset without removing the homologs. The column header `X\% training' indicates the fraction of the labeled data used for training and tuning the model with the rest (100-X)\% used as test data. We report the average AUC-PR over 10 random train-test splits (stratified splits that maintain the class-skew of 1:100). The standard deviation is also shown. The performance of the best baseline and the overall best method (BSL-MTL) is highlighted in bold.}
\label{resultsTable}
\begin{small}
\begin{center}
\def\arraystretch{1.2}
\begin{tabular}{c|ccccccc}
\toprule
& \multicolumn{3}{c}{\textbf{10\% training}} && \multicolumn{3}{c}{\textbf{30\% training}} \\ \cline{2-4}\cline{6-8}
& \textit{Ebola} & \textit{Hep-C} & \textit{Influenza} && \textit{Ebola} & \textit{Hep-C} & \textit{Influenza} \\ \midrule
Homolog & & & && & & \\
STL (Ridge Reg.)  & 0.189$\pm$.09 & 0.702$\pm$.08 & 0.286$\pm$.02 && 0.130$\pm$.03 & 0.802$\pm$.03 & 0.428$\pm$.03 \\ 
MMTL \citep{pontil04} & 0.113$\pm$.04 & 0.767$\pm$.03 & 0.321$\pm$.02 && 0.129$\pm$.02 & 0.802$\pm$.04 & 0.430$\pm$.03  \\ 
%Trace-norm  & 0.199$\pm$.11 & 0.767$\pm$.03 & 0.318$\pm$.02 && 0.207$\pm$.02 & 0.808$\pm$.02 & 0.409$\pm$.03 \\ 
Sparse+low-rank \citep{chen2012} & 0.144$\pm$.07 & 0.767$\pm$.02 & 0.318$\pm$.02 && 0.153$\pm$.02 & \textbf{0.814}$\pm$.01 & 0.414$\pm$.03  \\ 
%Dirty model \citep{jalali2010} & 0.074$\pm$.03 & 0.767$\pm$.04 & 0.324$\pm$.02 && 0.165$\pm$.02 & 0.813$\pm$.03 & 0.412$\pm$.03  \\ 
MTPL \citep{me_ismb_2013} & \textbf{0.217}$\pm$.08 & 0.695$\pm$.02 & 0.345$\pm$.02 && \textbf{0.260}$\pm$.05 & 0.713$\pm$.01 & \textbf{0.496}$\pm$.03 \\ 
IMC \citep{nagarajan} & 0.087$\pm$.04 & \textbf{0.779}$\pm$.02 & \textbf{0.362}$\pm$.01 && 0.122$\pm$.02 & 0.801$\pm$.01 & 0.410$\pm$.03  \\ \midrule
BSL-MTL (this work) & \textbf{0.233}$\pm$.10 & \textbf{0.807}$\pm$.02 & \textbf{0.486}$\pm$.02 && \textbf{0.361}$\pm$.03 & \textbf{0.842}$\pm$.01 & \textbf{0.560}$\pm$.02  \\ \bottomrule
\end{tabular}
\end{center}
\end{small}
\end{table*}

@@

>>>>

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

%\usepackage[top=1.3in, bottom=1.3in, left=1in, right=1in]{geometry}
\usepackage{floatrow}
\usepackage{graphicx} % more modern
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{subfigure} 
\usepackage{natbib}
%\usepackage[footnotesize]{caption}
\usepackage{multirow}
\usepackage{url}

@@

\newfloatcommand{capbtabbox}{table}[][11cm]
\newcommand{\argmin}{\operatornamewithlimits{argmin}}

%%%%%%%%%%%%%%%%%%%%%%%%%%
%\setcounter{tocdepth}{3}
%\setcounter{secnumdepth}{0}
%%%%%%%%%%%%%%%%%%%%%%%%%%

%\access{Advance Access Publication Date: Day Month Year}
%\appnotes{Manuscript Category}

\begin{document}
\firstpage{1}

%\subtitle{Subject Section}

\title[Multitask matrix completion]{Multitask matrix completion for learning protein interactions across diseases}

\author[Kshirsagar \textit{et~al}.]{Meghana Kshirsagar$^{1}$, Keerthiram Murugesan$^{2}$, Jaime G. Carbonell$^{2}$ and Judith Klein-Seetharaman$^{3}$}
\address{$^{1}$Memorial Sloan Kettering Cancer Center, 1275 York Ave., NY 10021\\
$^{2}$Language Technologies Institute, Carnegie Mellon Univ., 5000 Forbes Ave., Pittsburgh PA 15213, USA\\
$^{3}$Metabolic \& Vascular Health, Warwick Medical School, Univ. of Warwick, Coventry, UK
}

%\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

%\abstract{\textbf{Motivation:}
%Disease causing pathogens such as viruses, introduce their proteins into the host cells 
%where they interact with the host's proteins enabling the virus to replicate inside the host.
%These interactions between pathogen and host proteins are key to understanding infectious diseases.
%Often multiple diseases involve phylogenetically related or biologically similar pathogens. Here we present a multitask learning method to jointly model interactions between human proteins and three different, but related viruses: \textit{Hepatitis C}, \textit{Ebola} virus and \textit{Influenza A}. \\
%\textbf{Results:}
%Our multitask matrix completion based model uses a shared low-rank structure in addition to a task-specific sparse structure to incorporate the various interactions.
%We obtain between 7 to 39 percentage points improvement in predictive performance over prior state-of-the-art models. 
%We show how our model's parameters can be interpreted to reveal both general and specific
%interaction-relevant characteristics of the viruses. \\
%\textbf{Availability:}
%Our code and data is available at: \url{http://www.cs.cmu.edu/~mkshirsa/bsl_mtl.tgz}\\
%\textbf{Contact:}
%\href{mkshirsa@cs.cmu.edu}{mkshirsa@cs.cmu.edu}
%}

\maketitle

\begin{abstract}
\section{Motivation}
Disease causing pathogens such as viruses, introduce their proteins into the host cells 
where they interact with the host's proteins enabling the virus to replicate inside the host.
These interactions between pathogen and host proteins are key to understanding infectious diseases.
Often multiple diseases involve phylogenetically related or biologically similar pathogens. Here we present a multitask learning method to jointly model interactions between human proteins and three different, but related viruses: \textit{Hepatitis C}, \textit{Ebola} virus and \textit{Influenza A}. 
\section{Results}
Our multitask matrix completion based model uses a shared low-rank structure in addition to a task-specific sparse structure to incorporate the various interactions.
We obtain between 7 to 39 percentage points improvement in predictive performance over prior state-of-the-art models. 
We show how our model's parameters can be interpreted to reveal both general and specific
interaction-relevant characteristics of the viruses. \\
\section{Availability}
Our code and data is available at: \url{http://www.cs.cmu.edu/~mkshirsa/bsl_mtl.tgz}\\
\section{Contact}
\href{mkshirsa@cs.cmu.edu}{mkshirsa@cs.cmu.edu}
\end{abstract}

@@

This leads us to the following model that incorporates the above ideas. The interactions matrix $M_t$ of task $t$ can be written as: $M_t = \mu_t *\; \textrm{(shared component)} + (1-\mu_t) *\; \textrm{(specific component)} $, with hyperparameter $\mu_t$ allowing each task to customize its amount of shared and specific components.

To incorporate the above ideas, we assume that the interactions matrix $M$ is generated from two components. 
The first component has low-rank latent factors over the human and virus proteins,
with these latent factors jointly learned over all tasks. The second component involves a task specific parameter, on which we
additionally impose a sparsity constraint as we do not want this parameter to overfit the data. Section~\ref{ourmodel} discusses
our model in detail. We trade-off the relative importance of the two components using task-specific hyperparameters.
Our model can thus learn what is conserved and what is different across pathogens, rather than having to specify it manually.

@@

\noindent\textbf{Prediction of host-pathogen PPIs}: The most reliable experimental methods for studying protein-protein interactions (PPI) are often very time-consuming and expensive, making it hard to investigate the prohibitively large set of possible host-pathogen interactions -- for example, the bacterium \textit{Bacillus anthracis} which causes anthrax has about 2321 proteins which when coupled with the 100,000 or so human proteins gives $\approx$232 million protein pairs to test, experimentally. Computational techniques complement laboratory-based methods by predicting highly probable PPIs. These techniques use the known interactions data from previous experiments and predict the most plausible new interactions. In particular, supervised machine learning based methods use the few known interactions as training data and formulate the interaction prediction problem in a classification setting, with target classes: ``interacting'' or ``non-interacting''. Features are derived using various attributes of the two proteins such as: protein sequences from UniProt \cite{UniProt}, protein structure from PDB, gene ontology from GO database \cite{GO}.

@@

\begin{figure*}
\begin{floatrow}
\ffigbox[0.28\textwidth]{%
%trim option's parameter order: left bottom right top
\includegraphics[scale=0.35, trim = 0.2 9.2cm 4cm 0]{fig1.eps}
}
{
\caption{Multiple bipartite graphs: on the left are host proteins and on the right are proteins from different virus species. Edges represent protein interactions.}
\label{fig:hpppigraph}
}
\capbtabbox[0.68\textwidth]{%
\begin{footnotesize}
\begin{tabular}{c|r|r|r} 
\hline
Task $\rightarrow$ &{\textit{Influenza A} } &\textit{Hepatitis C} & {\textit{Ebola}}\\
& (28 strains) & (5 strains, many isolates)  & (3 strains) \\
number of HP PPIs (positives) & 848 & 981 & 90 \\\hline
number of non-homologous PPIs & 274 & 270 & 75 \\ \hline
\# of unique virus proteins in dataset & 54 & 151 & 2 \\\hline
\# of unique human proteins in dataset & 362 & 385 & 88 \\\hline
total \# of virus proteins in & 542 & 163 & 38 \\
UniProtKB across strains $^\dagger$ & & & \\ \hline
number of negatives & 84800 & 98100 & 9000 \\\hline
density of observed graph$^\ddagger$ (as \%) & .15 & .60 & .20 \\\hline
\multicolumn{4}{l}{\scriptsize{HP PPI: host-pathogen protein protein interactions}}\\
\multicolumn{4}{l}{\scriptsize{$\ddagger$ Density = (no. of positives + no. of negatives)/ total number of possible edges. The denominator was computed }} \\
\multicolumn{4}{l}{\scriptsize{\hspace{0.1cm} considering all proteins from all strains of the virus and $\approx$ 100,000 human proteins.}}\\
\multicolumn{4}{l}{\scriptsize{$\dagger$ this is the number of `reviewed' proteins on UniProtKB from all strains in our data}} \\ \hline
\end{tabular}
\end{footnotesize}
}
{
\caption{Tasks and their sizes (a column corresponds to one bipartite graph between human proteins and the virus indicated in the column header). More importantly, each task represents several strains of one virus species. We pool together PPI from all strains as the number of interactions per strain is very small and not representative. PPI data from different strains often involves homologous proteins (virus proteins or/and human proteins) as these were curated from different publications/studies involving the virus. The second row gives a conservative estimate of the number of non-homologous PPI. These were obtained by computing BLAST sequence similarity (see Section \ref{sec:phisto} for details). All pathogens are single stranded RNA viruses. The last row shows that each of our graphs is extremely sparse.} 
\label{datasets}
}
\end{floatrow}
\end{figure*}

@@

As before, the shared factors $U$ and $V$ are both $\mathbb{R}^{d_t \times k}$ (where the common dimensionality $d_t$ of the two node types is assumed for convenience). The matrix $S_t \in \mathbb{R}^{d_t \times d_t}$ is a sparse matrix. The objective function for the multitask model is given by:
\begin{equation}
\label{mtl_objective}
\begin{array}{ll}
\mathcal{L}(U,V,\{S_t\}) & = \displaystyle{ \frac{1}{N} \sum_{t=1}^T \sum_{(i,j) \in \Omega_t}} c^t_{ij} \; \ell \big( M_{t_{ij}}, \mathbf{x}_{ti}^\intercal H_t \mathbf{y}_{tj} \big)^2 + \\
& \lambda ( \|U\|^2_F + \|V\|^2_F ) + \displaystyle{\sum_{t=1}^T \sigma_t \|S_t\|_1} 
\end{array}
\end{equation}

@@

\begin{algorithm}[!t]   \caption{Alternating Least Squares}
   \label{alg}
   %\begin{minipage}{0.9\textwidth}
   \begin{small}
\begin{algorithmic}[1]
\STATE \textbf{Input}: \\
$k$ : number of latent factors\\
$\Gamma$: pairs of entities for initialization\\
For every task $t$,\\
${\{\mathbf{x}_{ti}\}, \{\mathbf{y}_{tj}\}}$: feature vectors for pathogen and host proteins resp.\\
$\Omega_t$: the observed entries of the matrix $M_t$\\
\STATE \textbf{Initialization}: \\
\STATE An iteration $r$, let $\mathbf{S}^r$ represent $\{S^r_t\}_{t=1}^T$
\STATE $\mathbf{S}^0 \leftarrow 0$
\STATE $U^0 \leftarrow$ top-$k$ left singular vectors and $V^0 \leftarrow$ top-$k$ right singular vectors from the SVD of $\displaystyle{\sum_{(i,j) \in \Gamma}} \mathbf{x}_i \mathbf{y}_j^\intercal$\\
\STATE $\mathcal{L}^0$ : initial loss
\REPEAT
  \STATE $U^{r+1} \leftarrow \displaystyle{\argmin_{U}} \; \mathcal{L}(U, V^r, \mathbf{S}^r)$  \STATE $V^{r+1} \leftarrow \displaystyle{\argmin_{V}} \; \mathcal{L}(U^{r+1}, V, \mathbf{S}^r) $  \STATE For each task $t$\\
  \hspace{0.5cm} $S^{r+1}_t \leftarrow \displaystyle{\argmin_{S_t}} \; \mathcal{L}(U^{r+1}, V^{r+1}, \mathbf{S}^r_{-t}) $
  \STATE Compute $\mathcal{L}^{r+1}$ and let $\delta \leftarrow (\mathcal{L}^r - \mathcal{L}^{r+1})/\mathcal{L}^r$
\UNTIL {$\delta < \tau$}%\While
\end{algorithmic}\end{small}
%\end{minipage}
\end{algorithm}

\noindent \textbf{Optimization}:
The function $\mathcal{L}(U, V, \{S_t\})$ is non-convex. However, it is convex in every one of the parameters (i.e when the other parameters are fixed) 
and a block coordinate descent method called alternating least squares (ALS) is commonly used to optimize such functions. 
To speed up convergence we use an adaptive step size. The detailed optimization procedure is shown in Algorithm \ref{alg}.

@@

\section{Dataset}
\label{sec:datasets}
We use three human-virus PPI datasets from the PHISTO \citep{phisto} database (version from 2014), the characteristics of which are summarized in Table \ref{datasets}. 
The \textit{Influenza A} task includes various strains of flu such as: H1N1 (Puerto Rico, Texas strains etc), H3N2 (England strain etc), H5N1, H7N3.
Similarly, the \textit{Hepatitis} task includes strains such as of the virus. 
All three are single-strand RNA viruses, with \textit{Hepatitis} being a positive-strand ssRNA whereas \textit{Influenza} and \textit{Ebola} are negative-strand viruses. The density of the known interactions is quite small when considering the entire proteome (i.e all known proteins) of the host and pathogen species (row-4 in Table \ref{datasets}). \\

\subsection{Addressing the homologs in PHISTO}
\label{sec:phisto}
The interactions data reported in PHISTO database is aggregated from several host-pathogen interactions data sources such as Virus-Mint, PHI-base etc. We found that many of the interactions (within a task) involve homologous PPI - virus protein homologs from two strains interacting with either the same human protein or with homologous human proteins. Such PPI usually come from different publications/ studies, but they introduce a lot of redundancy in the dataset. Removing such PPI is challenging as there are several criteria to consider, the most important being: \textit{what threshold of sequence similarity constitutes a homolog}? The very existence of sequence similarity between proteins is what makes it possible for biologists to predict unknown interactions and most computational methods rely on it for PPI prediction.

In row-2 of Table \ref{datasets}, we give a conservative estimate of the number of non-homologous PPI. For this, we identify homologous PPI and retain only one of them in the dataset. Homologs were obtained using BLAST sequence alignment (blastp) using an e-value cut-off threshold of $1e^{-10}$. This is a relaxed cut-off since there are cases where two functionally different proteins (either virus-virus or human-human) with low query coverage and low identity end up as `homologs'. For example, human proteins Q8WV28 (gene BLNK, B-cell linker protein) and O00459 (gene PIK3R2, Phosphatidylinositol 3-kinase) are considered homologs by this cut-off. The number of non-homologous PPI should therefore be considered a lower limit.

In our experiments section, we thus evaluate our models in two settings - (1) on the original dataset (2) by removing homologous PPI.

%and thus interologs (within a task), the presence of which may not give a clear picture of a computational model. However these are reported by different publications/studies and were found using different experimental methods, so we do not exclude them and instead report performance on

\noindent\textbf{Features}: 
Since the sequence of a protein determines its structure and consequently its function, it may be possible to predict PPIs using the amino acid sequence of a protein pair. \cite{shen:2007} introduced the ``conjoint triad model'' for predicting PPIs using only amino acid sequences. They partitioned the twenty amino acids into seven classes based on their electrostatic and water affinities.\footnote{For details of these classes, please refer to the supplementary or the original paper}
A protein's amino acid sequence is first transformed to a class-sequence (by replacing each amino acid by its class).
For $k$=3, they count the number of times each distinct tri-mer (set of three consecutive amino acids) occurred in the sequence.
Since there are 343 ($7^3$) possible tri-mers (with an alphabet of size 7), the feature vector containing the tri-mer frequency counts will have 343 elements.
To account for protein size, they normalized the counts by linearly transforming them to lie between 0 and 1.
Thus the value of each feature in the feature vector is the normalized count for each of the possible amino acid three-mers. We use di-, tri- and four-mers thus leading to a total of 2793 features ($7^2 + 7^3 + 7^4$).
Such features have been successfully applied in prior work \citep{dyer07,me_ismb_2013}.

\section{Experimental setup}

@@

\subsection{Comparison with a homolog baseline}
In addition to the machine learning methods presented in Section \ref{mlmethods}, we also tried a simple homology based approach that is based on protein sequence identity. We use the entire human-virus interactome from PHISTO as input to this method. Given a human protein $h$, we rank all virus proteins $v$ based on the following simple scheme. let $V(h)$ be a set of all virus proteins interacting with the human protein $h$ and let $H(v)$ be all human proteins interacting with the virus protein $v$. We score each pair $(h, v)$ from the bipartite graph as follows: $\max \{ \max $ sequence identity between $h$ and all host proteins in $H(v)$; $\max $ sequence identity between $ v $ and all virus proteins in $V(h) \}$. This baseline is shown in the first row as \textit{Homolog} in Table \ref{resultsTable}.

@@

\begin{table*}[t]\caption{Area Under the Precision-Recall curve for each task on the PHISTO dataset without removing the homologs. The column header `X\% training' indicates the fraction of the labeled data used for training and tuning the model with the rest (100-X)\% used as test data. We report the average AUC-PR over 10 random train-test splits (stratified splits that maintain the class-skew of 1:100). The standard deviation is also shown. The performance of the best baseline and the overall best method (BSL-MTL) is highlighted in bold. The Homolog baseline performs comparably to our BSL-MTL method on the \textit{Ebola} task. This is explained by the virus-human PPI graph, from several other viruses, that the Homolog baseline has access to.}
\label{resultsTable}
\begin{small}
\begin{center}
\def\arraystretch{1.2}
\begin{tabular}{c|ccccccc}
\toprule
& \multicolumn{3}{c}{\textbf{10\% training, 90\% test}} && \multicolumn{3}{c}{\textbf{30\% training, 70\% test}} \\ \cline{2-4}\cline{6-8}
& \textit{Ebola} & \textit{Hep-C} & \textit{Influenza} && \textit{Ebola} & \textit{Hep-C} & \textit{Influenza} \\ \midrule
Homolog & \textbf{0.230}$\pm$.06 & 0.178$\pm$.01 & 0.158$\pm$.01 && \textbf{0.311}$\pm$.03 & 0.180$\pm$.01 & 0.198$\pm$.01 \\
STL (Ridge Reg.)  & 0.189$\pm$.09 & 0.702$\pm$.08 & 0.286$\pm$.02 && 0.130$\pm$.03 & 0.802$\pm$.03 & 0.428$\pm$.03 \\ 
MMTL \citep{pontil04} & 0.113$\pm$.04 & 0.767$\pm$.03 & 0.321$\pm$.02 && 0.129$\pm$.02 & 0.802$\pm$.04 & 0.430$\pm$.03  \\ 
%Trace-norm  & 0.199$\pm$.11 & 0.767$\pm$.03 & 0.318$\pm$.02 && 0.207$\pm$.02 & 0.808$\pm$.02 & 0.409$\pm$.03 \\ 
Sparse+low-rank \citep{chen2012} & 0.144$\pm$.07 & 0.767$\pm$.02 & 0.318$\pm$.02 && 0.153$\pm$.02 & \textbf{0.814}$\pm$.01 & 0.414$\pm$.03  \\ 
%Dirty model \citep{jalali2010} & 0.074$\pm$.03 & 0.767$\pm$.04 & 0.324$\pm$.02 && 0.165$\pm$.02 & 0.813$\pm$.03 & 0.412$\pm$.03  \\ 
MTPL \citep{me_ismb_2013} & 0.217$\pm$.08 & 0.695$\pm$.02 & 0.345$\pm$.02 && 0.260$\pm$.05 & 0.713$\pm$.01 & \textbf{0.496}$\pm$.03 \\ 
IMC \citep{nagarajan} & 0.087$\pm$.04 & \textbf{0.779}$\pm$.02 & \textbf{0.362}$\pm$.01 && 0.122$\pm$.02 & 0.801$\pm$.01 & 0.410$\pm$.03  \\ \midrule
BSL-MTL (this work) & \textbf{0.233}$\pm$.10 & \textbf{0.807}$\pm$.02 & \textbf{0.486}$\pm$.02 && \textbf{0.361}$\pm$.03 & \textbf{0.842}$\pm$.01 & \textbf{0.560}$\pm$.02  \\ \bottomrule
\end{tabular}
\end{center}
\end{small}
\end{table*}

@@

\begin{figure}
%trim option's parameter order: left bottom right top
\includegraphics[scale=0.165, trim = 0 0.8cm 0 0]{fig4.eps}
\includegraphics[scale=0.165, trim = 0 0 0 0]{fig8.eps}
\caption{(\textit{Left}) Top tri-mer sequence motifs from virus proteins across all three viruses. These correspond to features important for interactions and is derived from our model parameter $H = U V^{T}$. (\textit{Right}) Motif corresponding to enriched 4-mer patterns found by our model (parameter: $S_{hepc}$) for the Hepatitis-C virus task.}%
\label{fig:virus_motifs}
\end{figure}
